Pythran: Bringing OpenMP to a Python Subset

### 2013-01-29 doc/papers/europar2013/paper.tex
Draft version.

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\begin{document}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies the
existing approaches and shows a critical lack of parallelism support.
Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives to the
Python language in order to add fine-grain parallelism support to Python in the
context of a Python subset called Pythran. Static compilation of this language
into C++ and the underlying runtime is described in
Section~\ref{sec:python-static}.
Section~\ref{sec:python-optim} presents several optimizations applied by the compiler to
optimized generated code with respect to parallelism. The Pythran compiler is
benchmarked on several scientific applications and compared to both interpreted
and native code in Section~\ref{sec:validation}.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shedskin~\cite{shedskin2006}, a Python to C++ compiler also are viable way to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library
proposes.~\footnote{Support for STM in PyPy is a work in
progress.}~\footnote{Shedskin is compatible with several Python modules that
provide coarse-grained, fork-based parallelism.}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++.

@@

\subsubsection{Variable Scope}

\subsubsection{Iterators}

@@

@@

\cite{wang2012}

@@

@@

\subsubsection{Shared References}

\subsubsection{Iterator Category}

%=========================================================
\section{Optimizations for Parallelism}\label{sec:python-optim}
%=========================================================

\subsection{Limiting Copies}

@@

@@

\subsection{Comparison with CPython}

\subsection{Comparison with Cython}

@@

>>>>

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\newcommand\see{\emph{cf.\ }}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}

\lstset{language=python}

\begin{document}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

@@

This hybrid approach, based on a mix of interpreted and native code, is getting
wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies
the existing approaches and shows a critical lack of parallelism support.
Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
to the Python language in order to add fine-grain parallelism support to Python
in the context of a Python subset called Pythran. Static compilation of this
language into C++ and the underlying runtime is described in
Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
scientific applications and compared to both interpreted and native code in
Section~\ref{sec:validation}.

@@

Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
Shedskin~\cite{shedskin2006}, a Python to C++ compiler also are viable way to
turn regular Python codes into optimized ones, but they do not offer support for
fine-grained parallelism beyond what the standard library
proposes.
%~\footnote{Shedskin is compatible with several Python modules that
%provide coarse-grained, fork-based parallelism.}

@@

OpenMP is a standard API for parallel programming for Fortran, C and C++. It
consists in a set of parallelizing directives and a few runtime library calls.
If OpenMP is not activated, the directives are ignored, thus enabling
incremental parallelization of the original source code while keeping the
original code structure.

The languages targeted by OpenMP are statically compiled. This section studies
the semantic adaptation required to use the same API for a scripting language
such as Python.

@@

OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
on structured blocks, represented by an instruction inC/C++ and delimited by
comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
attached to a specific instruction.

While Pythran does have a decorator mechanism~\footnote{\see PEP 319,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation fo
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to random instructions.
PLW~\cite{dongara2007} suggests to use string instructions to hold such
decorations. As Python does not have anonymous block~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.}, one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Alternatively, the syntax
\texttt{if 'my annotation':} is also supported.
Listing~\ref{lst:instruction-annotation} shows how to annotate a parallel loop
with reduction this way.

\begin{figure}
    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
    loop with reduction annotated with OpenMP directive.}]
'omp parallel for reduction(+:r)'
for auto in my_container:
  r+=x
  \end{lstlisting}
\end{figure}

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope. Consequently, all variables referred to in a function are to be
considered when building such variable lists: there is no concept of variable
local to a block. Using the \texttt{default(none)} clause is possible to ensure
that no variable gets forgotten when building such lists, as a compile-time
error is raised if a variable is referred in a block but not in the associated
parallel region.

The \texttt{reduction(\emph{operator}: \emph{list}} is used to handle data
dependencies when performing a parallel reduction. The supported operator
depends on the input and backend language: Python has \texttt{min}/\texttt{max} operators but
C/C++ have not. Python does not have \texttt{\&\&} or \texttt{||}
operators~\footnote{Although similar, the \texttt{and} and \texttt{or} keywords
have a special meaning in Python.} while C/C++ have.

\subsubsection{Parallel For and Iterators}

The core directive to handle data parallelism is the \texttt{for} directive that
distribute the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order, that is it is possible
to increase the iterator by any integer and to decide which iterator is the
greatest among two. The integer used as loop indices in Fortran and C satisfy
this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
trait. But a Python iterator only advance by a step of one until it is
exhausted, in which case it raises an exception. They only have forward iterator
behavior.

The extension of Python iterator to random access iterators is direct for the
standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
\texttt{vector}s.}, \texttt{set} or \texttt{dict}.

Python also generalizes the concept of iterators through generators, which also
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}, is
easily extended to support random access, but it is generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

A particular class of generator behaves like adaptor, in the sense that they
apply a particular expression on each value of the iterator. It is the case of
the \texttt{enumerate} builtin that yields each element of the enumerated
iterator associated with its index. Generator expressions such as \texttt{(x*x
for x in l)} have the same behavior. These generators are random iterators only
if their input iterator is a random iterator itself.

Finally, if the iterator yields a tuple, it is possible to unpack it inside the
\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
all the unpacked variables are considered as iterators, especially with respect
to default privatization rules. In the given example, \texttt{i} and \texttt{v}
are private, and the parallel iteration is valid because the input of
\texttt{enumerate} is a list, which allows random access iteration.

\begin{figure}
    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
    loop with tuple unpacking.}]
'omp parallel for'
for i,v in enumerate([2, 3, 5, 7, 11]):
  print i, ':', v
  \end{lstlisting}
\end{figure}

@@

The OpenMP runtime library makes it possible to retrieve the active thread id,
whether dynamic scheduling is enabled and so on. All the functions are declared
in the \texttt{<omp.h>} header, and have a default behavior when OpenMP is not
activated.

Providing a biding to these library in the Python does not raise
particular problems, as the signature of these functions only involves passing
integers as parameters or return values, except for the mutex manipulation. In
that case an opaque type is used to represent the native type.

The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
activated, and can be used for portability with system where OpenMP is not
available. Python does not have a preprocessor, but is possible to catch the
import exception if the \texttt{omp} module is not found.

@@

A validation suite for OpenMP has been proposed in~\cite{wang2012} for C and
Fortran. We have ported it to Python, and we also extended it to validate the
corner cases specific to Python described in this section. An example of such
test case is given in Listing~\ref{lst:openmp-validation}.

\begin{figure}
    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
    OpenMP validation test case.}]
def omp_parallel_for_if():
  import omp
  using = num_threads = 0
  sum = sum2 = 0
  LOOPCOUNT=1000
  'omp parallel for private(i) if(using == 1)'
  for i in range(LOOPCOUNT+1):
    num_threads = omp.get_num_threads()
    sum+=i
  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
  return known_sum == sum and num_threads == 1
    \end{lstlisting}
\end{figure}

We have used the Pythran tool described in the following section to turn each
Python test function into a C++ function with the same directives and runtime
calls. Apart from the \texttt{threadprivate} directive and the
\texttt{collapse(\emph{n}} clause, all tests were successful.
\texttt{threadprivate} directives were held by global variables not supported in
Pythran~; and the C++ code generated by Pythran does not preserve the perfect
loop nesting required by the \texttt{collapse} clause.

@@

Pythran is subset of Python focused on scientific computing. It is implicitly
statically typed and support most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard module are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
partially supported. The associated compiler turns Pythran code, eventually
annotated with OpenMP and exported function type annotation, into C++ code. Its
processing is summarized in Figure~\ref{fig:pythran-compiler}. The purpose of
this section is not to dig into the internal of the compiler, but rather to
focus on the impact of the parallelism layer, especially on the runtime library,
\texttt{pythonic}.

\begin{figure}

\centering
\begin{tikzpicture}[
file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center},
    tool/.style={draw=black!50,fill=black!10,circle, align=center}]
    \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
\node[file] (cxx) [below=of pythran] {\textbf{C++}};
\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

\draw[very thick, ->] (python) -- (pythran);
\draw[very thick,dashed, ->] (annotation) -- (python);
\draw[very thick,dashed, ->] (omp) -- (python);
\draw[very thick, ->] (pythran) -- (cxx);
\draw[very thick, ->] (cxx) -- (gxx);
\draw[very thick, ->] (boost) -- (gxx);
\draw[very thick, ->] (pythonic) -- (gxx);
\draw[very thick, ->] (gxx) -- (so);
\end{tikzpicture}
\caption{Pythran compiler workflow.}
\label{fig:pythran-compiler}
\end{figure}

The standard library of C++11, unlike its C++03 version, is thread-safe. Pythran
runtime is based on the STL and provides the same guaranty, so that the data
dependences in the original code remains explicit.

A critical point when designing the pythonic runtime was memory management, the
very same aspect that lead to the use of a GIL in CPython. Pythran handles the
problem by refusing recursive types, which makes it possible to solely rely on
reference counting for memory management. It can be implemented through the
thread-safe shared pointer mechanism provided in C++11 standard library.

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. It limits parallelism, so reducing the
number of copies remains an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing may be a trouble. To avoid a reference increment, one can use pass
parameters by reference. Using an interprocedural memory effect analysis, Pythran
determines for each argument whether it must be passed using reference or const
reference to prevent this extra cost.

@@

During Python to C++ translation, Pythran adopts a blind strategy: it does not
understand the semantic of the annotations. Instead, it just split each
annotation into a context-sensitive part ---the variable names--- and a
context-insensitive part ---the clauses---, and attach them to the proper
instruction in the AST. It then ensures a bijective translation between Python
instruction and generated C++ instruction so that the OpenMP directive is
regenerated on the proper instruction. The same approach is used at the
expression level, to be compatible with the \texttt{if} clause.

@@

When passing containers from Python to C back and forth, a copy of the whole
data is made to turn the type agnostic, non-contiguous Python data into dense
typed ones. This extra copy implies an extra cost that is not negligible: the
time taken to pass two lists of float from Python to C++ is only twice less than
the time taken to compute the dot product of the same lists directly in Python.
Following Amdahl's law, this transfer cost greatly hinders parallelization, and
is a well known issue.

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. Numpy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and Numpy
packages. Basically, a numpy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. As a tool for scientific computing, Pythran supports such a
structure.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, we
successively parallelized it with OpenMP, turned it into Python with a Pythran
compatible kernel and parallelized the Python version using the approach
described in this paper. We also measurer the SLOC for the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It
non only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

\begin{table}

    \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
        \hline
        C       & 102   & 25 & 22 \\
        Python  & 30    & 676 & --\\
        Pythran & 30    & 38 &  22 \\
        Pythran+OMP    & 30    & 11 & no OpenMP support\\
        \hline
    \end{tabular}

    \caption{Comparison of several versions of the Hyantes prototype.}
    \label{tbl:hyantes}

\end{table}

We then compare Pythran with Cython. The approach share some similarities: they
both generate code written in a rather low level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information that Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They both translate explicit fine-grained
parallelism through code annotations or specific python constructs,
respectively.

Pythran exposes the full OpenMP interface to the user, thus enabling both data
and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
not the case in Cython, where:

\begin{itemize}
    \item Only loops can be made parallel, through the use of a dedicated
        intrinsic \texttt{prange} as loop iterator. To use more advanced
        parallelism construction, one must rely on the possibility to write raw C
        code embedded in cython.~\todo{SG->PB: tu m'explique ce truc de C dans
        cython ?}

    \item Cython automatically infers reductions and variable privacy, giving
        less control to the end-user.

    \item As Cython allows the mix of Python and Cython code, the user is
        responsible from releasing the GIL.

    \item The code included in a parallel region interacts with the Python code
        in a restricted manner. For instance, it is impossible to use a function
        imported from a Python module, while it is still possible to use native
        C functions.
\end{itemize}

Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrates the
difference between the two approaches. It shows that while being more integrated
in Python, the Cython approach is more verbose.

\begin{figure}[ht]

    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
import math
cimport cython
from cython.parallel import parallel, prange
from libc.math cimport sqrt
def sum_sqrt():
    cdef float r=0
    cdef int i
    with nogil, parallel():
        for i in prange(10000000):
            r+= sqrt(i)
    return r
\end{lstlisting}
\end{figure}
%
\begin{figure}[ht]
    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython Impelmentation
    of a Parallel Reduction.}]
#pythran export sum_sqrt()
import math
def sum_sqrt():
    r=0
    "omp parallel for private(i) reduction(+:r)"
    for i in xrange(10000000):
        r+= math.sqrt(i)
    return r
    \end{lstlisting}
\end{figure}

The performance of the two approaches is illustrated in
Figure~\ref{fig:cython-pythran}.

\begin{figure}[ht]
    \includegraphics[width=\textwidth]{cython}
    \caption{Comparison of Cython en Pythran generated code performance.}
    \label{fig:cython-pythran}
\end{figure}

The benchmarked codes are typical mathematical, image-processing or linear
algebra kernels. All these kernels have been written in Cython and Python
---compatible with Pythran--- and annotated through the mechanism of each
language to exhibit parallelism, then compiled into native code using GCC 4.1
and the \texttt{-O2 -fopenmp} flag. Their execution time when called from the
Python interpreter is measured through the \texttt{timeit} module. All results
are normalized to Cython sequential execution time.

This figure shows our inability to select proper benchmarking code, and to run
them on a real multiprocessor machine, but this is gonna change!~\todo{SG->PB:
c'est pour toi !}

@@

This paper studies the relevancy of adding OpenMP annotations to Python code. It
shows that providing minor semantic adaptations, these annotations are relevant
and makes it possible for Python code to benefit from multi-core performance
boost without worrying about the Global Interpreter Lock.

To achieve this goal, a translator from a subset of the Python language to C++
has been designed. This translator can turn regular Python modules
annotated with OpenMP directives and a few type annotations into native parallel
module. Both the input module and the generated one remain compatible with the
standard interpreter.

The approach is compared with Cython, an extension of the Python language used
to generate native module with an hybrid Python-C syntax that also provide means
to exhibits fine grained parallelism. Modules generated by Pythran outperform
modules generated by Cython for various kernels and both benefit from the
parallel boost.

Future works include the extension of the approach to OpenMP 4 and the
\texttt{target} clause that should make it possible to target MIC from Python.
There is also many vectorization possibility in Python, especially in the list
comprehension construction, that needs to be explored.

%=========================================================
\section*{Acknowledgments}
%=========================================================

This work has received fundings from the \texttt{Silkan} company and the French
ANR through the CARP project. The authors would like to thanks Adrien
\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime.
Fabien \textsc{Dagnat} and Mehdi \textsc{Amini} also contributed to this paper
through their many valuable advices.
